ANGSD-wrapper: utilities for analyzing next generation sequencing data

### 2016-05-02 manuscript.tex
jri minor edits

\begin{abstract}
High throughput sequencing has changed many aspects of population genetics, molecular ecology, and related fields, affecting both experimental design and data analysis. 
The software package ANGSD allows users to perform a number of population genetic analyses on high-throughput sequencing data.
ANGSD uses probabilistic approaches to calculate genome-wide descriptive statistics. 
The package makes use of genotype likelihood estimates rather than SNP calls and is specifically designed to produce more accurate results for samples with low sequencing depth. 
ANGSD makes use of full genome data while handling a wide array of sampling and experimental designs. 
Here we present ANGSD-wrapper, a set of wrapper scripts that provide a user-friendly interface for running ANGSD and visualizing results.
ANGSD-wrapper supports multiple types of analyses including estimates of nucleotide sequence diversity and performing neutrality tests, principal component analysis, estimation of admixture proportions for individuals samples, and calculation of statistics that quantify recent introgression. 
ANGSD-wrapper also provides interactive graphing of ANGSD results to enhance data exploration.
We demonstrate the usefulness of ANGSD-wrapper by analyzing resequencing data from populations of wild and domesticated \textit{Zea}. 
ANGSD-wrapper is freely available from \url{https://github.com/mojaveazure/angsd-wrapper}.
\end{abstract}

@@

A number of tools have recently been published to estimate population genetic descriptive statistics using high throughput sequencing data \citep{garrigan2013popbam, purcell2007plink, danecek2011variant, hutter2006genome}. 
 \citet{korneliussen2014angsd} recently published the software package ANGSD, which enables users to flexibly perform a large number of common population genetic analyses, including estimation of diversity statistics, admixture analysis including Patterson's D statistic \citep{Durand:2011jd}, site frequency spectrum estimation \citep{pmid22911679}, and calculation of neutrality test statistics \citep{korneliussen_calculation_2013}. 
ANGSD works directly with the alignment formats produced from standard high throughput sequence analysis pipelines, which removes the need for the user to transform the data into a niche format. 
One of the most important features of ANGSD is that analyses are integrated over per-site genotype likelihoods, rather than only on pre-determined variable sites. 
This permits ANGSD to calculate common population genetic descriptive statistics on low-coverage sequencing data, and handle missingness due to variation in coverage. 
Further, this probabilistic framework is appealing to researchers studying non-model organisms where the lack of existing genomic resources or funding prevents the ability to obtain high quality data. 
Additionally, for many of these non-model organisms, existing tools are not able to take into account biological differences such as non-random mating into their statistical models.
Recent use of ANGSD in the study of non-model organisms such as the blue-eyed black lemur \citep{meyer2015evolutionary} and \textit{Ficedula} flycatchers \citep{burri2015linked} has allowed population genetic analysis despite using low coverage resequencing data.

@@

\section*{Methods}
ANGSD-wrapper is a set of configuration files and scripts written primarily in the Bash scripting language. 
The scripts can be run either on a standalone computer with a UNIX terminal, or on computing clusters where they can be submitted to a queuing system such as SGE \citep{Microsystems):2001:SGE:560889.792378}, Slurm \citep{Jette02slurm:simple} or TORQUE \citep{Staples:2006:TRM:1188455.1188464}. 
A Python installation \citep{Python} is required for some light, dynamic pre-processing of the data, and the statistical software R \citep{Rcitation} is required to make use of the visualization tools incorporated in ANGSD-wrapper.  
The visualization portion of ANGSD-wrapper also requires installation of the R packages \texttt{Shiny} \citep{shiny}, \texttt{APE} \citep{APE}, \texttt{Lattice} \citep{lattice}, \texttt{Hmisc} \citep{Hmisc}, \texttt{data.table} \citep{datatable}, \texttt{DT} \citep{javascriptDT}, and \texttt{shinythemes} \citep{shinythemes}, as well as \texttt{genomeIntervals} \citep{genomeIntervals} from Bioconductor \citep{bioconductor}.
These dependent packages are installed automatically upon first run of the visualization interface.

@@

Analyses supported by ANGSD-wrapper are shown in Table \ref{tab:methods}.
A detailed flowchart of each of these workflows is shown in Figure \ref{fig:supp2}, and additional details, documentation, a tutorial, and a wiki can be found on the GitHub page: \url{https://github.com/mojaveazure/angsd-wrapper/wiki}.

The visualization software included with ANGSD-wrapper is contained in a directory called `shinyGraphing.`
This application is started from within ANGSD-wrapper and launched locally from a standard web browser. 
This software provides a graphical user interface (GUI) to quickly and interactively plot results obtained from ANGSD-wrapper.  
Each tab in the GUI contains plots for different ANGSD methods.
In order to use the plotting software, the user navigates to the desired tab and uploads the appropriate results file. 
The Shiny server automatically parses ANGSD output files and creates the resulting plot(s) (Figure \ref{fig:theta}), which can be saved using the browser's built-in image saving capabilities.

@@

\section*{Results and Discussion}
As a demonstration of analyses in ANGSD-wrapper, we explore patterns of pairwise nucleotide diversity in a single genomic region of domesticated maize and wild teosinte. 
We used a subset of the resequenced samples from the Maize HapMap2 project (Table \ref{tab:samples}) and calculated summary statistics using a 10Mb region on chromosome 10 \citep{chia2012maize}. 
The data are available at \url{https://figshare.com/articles/Example_Data_tar_bz2/2063442} and can be downloaded from within ANGSD-wrapper using the command

@@

We first use the ``SFS'' method to estimate the site frequency spectrum (SFS) of both maize and its wild progenitor \textit{Zea mays} ssp. \textit{parviglumis}, assuming an inbreeding coefficient of $F=1$ for these highly inbred samples. 
The SFS and diversity statistics were calculated using ancestral states inferred by the ``Ancestral Sequence'' method from a single resequenced genome of \textit{Tripsacum dactyloides}.  
We show that the maize SFS is skewed towards more intermediate-frequency variants (Figure \ref{fig:figure3}A-B, Tajima's D of 0.2 and 0.0085 in maize and teosinte, respectively), likely a result of the bottleneck associated with maize domestication \citep{eyre1998investigation, Beissinger031666}.
Using the ``Thetas'' method we find further evidence of the domestication bottleneck, with mean levels of pairwise nucleotide diversity in this region in maize $\approx 25\%$ lower than in teosinte (0.0061 and 0.0082, respectively; Figure \ref{fig:figure3}C). 
Using the ``2DSFS method,'' which includes an \fst calculation, we find a mean \fst in this region of 0.116, nearly identical to the  genome-wide value of 0.11 reported in \cite{hufford2012comparative}.   
None of the genes in this region have been identified as potential domestication candidates \cite{hufford2012comparative}, consistent with the lack of extended regions of high \fst in our analysis (Figure \ref{fig:suppfst}).

@@

\section*{Conclusions}
ANGSD-wrapper provides an easy-to-use interface that simplifies many population genetic analyses implemented in ANGSD \citep{korneliussen2014angsd} and permits the exploration of genome-scale results through interactive visualization.
ANGSD-wrapper is under active development to incorporate updates to the ANGSD software package.  
\section*{Acknowledgments}
We acknowledge funding support from the US National Science Foundation (IOS-1238014 to JRI and IOS-1339393 to PLM), from a University of Minnesota Doctoral Dissertation Fellowship supporting TJYK, and from the University of California Davis Plant Sciences Department. We thank members of the Ross-Ibarra and Morrell Labs for discussion and software testing. We thank the authors of ANGSD and related programs for answering questions, particularly Matteo Fumagalli and Filipe Vieira. Finally, we would like to thank Felix Andrews for statistical advice, although we did not follow it. 
\clearpage
\singlespacing
\bibliographystyle{apa}
\bibliography{references}

>>>>

\begin{abstract}
High throughput sequencing has changed many aspects of population genetics, molecular ecology, and related fields, affecting both experimental design and data analysis. 
The software package ANGSD allows users to perform a number of population genetic analyses on high-throughput sequencing data.
Rather than require users to first call SNPs, however, ANGSD uses probabilistic approaches which can directly make use of genotype likelihoods, taking advantage of all of the sequencing data and producing more accurate results for samples with low sequencing depth
Here we present ANGSD-wrapper, a set of wrapper scripts that provide a user-friendly interface for running ANGSD and visualizing results.
ANGSD-wrapper supports multiple types of analyses including estimates of nucleotide sequence diversity and performing neutrality tests, principal component analysis, estimation of admixture proportions for individuals samples, and calculation of statistics that quantify recent introgression. 
ANGSD-wrapper also provides interactive graphing of ANGSD results to enhance data exploration.
We demonstrate the usefulness of ANGSD-wrapper by analyzing resequencing data from populations of wild and domesticated \textit{Zea}. 
ANGSD-wrapper is freely available from \url{https://github.com/mojaveazure/angsd-wrapper}.
\end{abstract}

@@

A number of tools have recently been published to estimate population genetic descriptive statistics using high throughput sequencing data \citep{garrigan2013popbam, purcell2007plink, danecek2011variant, hutter2006genome}. 
 \citet{korneliussen2014angsd} recently published the software package ANGSD, which enables users to flexibly perform a large number of common population genetic analyses, including estimation of diversity statistics, admixture analysis including Patterson's D statistic \citep{Durand:2011jd}, site frequency spectrum estimation \citep{pmid22911679}, and calculation of neutrality test statistics \citep{korneliussen_calculation_2013}. 
ANGSD works directly with alignment formats produced from standard high throughput sequence analysis pipelines, which removes the need for the user to transform the data into a software-specific format. 
One of the most important features of ANGSD is that analyses are integrated over per-site genotype likelihoods, rather than only on polymorphic sites identified \emph{a priori}. 
This permits ANGSD to calculate common population genetic descriptive statistics on low-coverage sequencing data, and handle missingness due to variation in coverage. 
Further, this probabilistic framework is appealing to researchers studying non-model organisms where the lack of existing genomic resources or funding prevents the ability to obtain high quality data. 
Additionally, for many of these non-model organisms, existing tools are not able to take into account biological differences such as non-random mating into their statistical models.
Recent use of ANGSD highlights these benefits, allowing population genetic analyses in non-model organisms such as the blue-eyed black lemur \citep{meyer2015evolutionary} and \textit{Ficedula} flycatchers \citep{burri2015linked} and inbred lines of wild and cultivated maize \citep{beissinger2015recent}.

@@

\section*{Methods}
ANGSD-wrapper is a set of configuration files and scripts written primarily in the Bash scripting language. 
The scripts can be run either on a standalone computer with a UNIX terminal, or on computing clusters where they can be submitted to a queuing system such as SGE \citep{Microsystems):2001:SGE:560889.792378}, Slurm \citep{Jette02slurm:simple} or TORQUE \citep{Staples:2006:TRM:1188455.1188464}. 
A Python installation \citep{Python} is required for some light, dynamic pre-processing of the data, and the statistical software R \citep{Rcitation} is required to make use of the visualization tools incorporated in ANGSD-wrapper.  
The visualization portion of ANGSD-wrapper also requires installation of the R packages \texttt{Shiny} \citep{shiny}, \texttt{APE} \citep{APE}, \texttt{Lattice} \citep{lattice}, \texttt{Hmisc} \citep{Hmisc}, \texttt{data.table} \citep{datatable}, \texttt{DT} \citep{javascriptDT}, and \texttt{shinythemes} \citep{shinythemes}, as well as \texttt{genomeIntervals} \citep{genomeIntervals} from Bioconductor \citep{bioconductor}; these are installed automatically upon first run of the visualization interface.

@@

Analyses supported by ANGSD-wrapper are shown in Table \ref{tab:methods}.
A detailed flowchart of each of these workflows is shown in Figure \ref{fig:supp2}, and additional  documentation, a tutorial, and a wiki can be found on the project GitHub page: \url{https://github.com/mojaveazure/angsd-wrapper/wiki}.

The visualization software included with ANGSD-wrapper is contained in a directory called `shinyGraphing.`
This application is started from within ANGSD-wrapper and launched locally from a standard web browser. 
This software provides a graphical user interface (GUI) to quickly and interactively plot results obtained from ANGSD-wrapper.  
Each tab in the GUI contains plots for different ANGSD methods.
In order to use the plotting software, the user navigates to the desired tab and uploads the appropriate results file. 
The Shiny server automatically parses standard ANGSD output files and creates the resulting plot(s) (Figure \ref{fig:theta}), which can be saved using the browser's built-in image saving capabilities.

@@

\section*{Results and Discussion}
As a demonstration of analyses in ANGSD-wrapper, we explore patterns of pairwise nucleotide diversity in a single genomic region of domesticated maize and wild teosinte. 
We used a subset of the resequenced samples from the Maize HapMap2 project \citep{chia2012maize} and calculated summary statistics using a 10Mb region on chromosome 10 (Table \ref{tab:samples}). 
The data are available at \url{https://figshare.com/articles/Example_Data_tar_bz2/2063442} and can be downloaded from within ANGSD-wrapper using the command

@@

We first use the ``SFS'' method to estimate the site frequency spectrum (SFS) of both maize and its wild progenitor \textit{Zea mays} ssp. \textit{parviglumis}, assuming an inbreeding coefficient of $F=1$ for these highly inbred samples. 
The SFS and diversity statistics were calculated using ancestral states inferred by the ``Ancestral Sequence'' method from a single resequenced genome of \textit{Tripsacum dactyloides}.  
We show that the maize SFS is skewed towards more intermediate-frequency variants (Figure \ref{fig:figure3}A-B, Tajima's D of 0.2 and 0.0085 in maize and teosinte, respectively), likely a result of the bottleneck associated with maize domestication \citep{eyre1998investigation, beissinger2015recent}.
Using the ``Thetas'' method we find further evidence of the domestication bottleneck, with mean levels of pairwise nucleotide diversity in this region in maize $\approx 25\%$ lower than in teosinte (0.0061 and 0.0082, respectively; Figure \ref{fig:figure3}C). 
Using the ``2DSFS method,'' which includes an \fst calculation, we find a mean \fst in this region of 0.116, nearly identical to the  genome-wide value of 0.11 reported in \cite{hufford2012comparative}.   
None of the genes in this region have been identified as potential domestication candidates \cite{hufford2012comparative}, consistent with the lack of extended regions of high \fst in our analysis (Figure \ref{fig:suppfst}).

@@

\section*{Conclusions}
ANGSD-wrapper provides an easy-to-use interface that simplifies many population genetic analyses implemented in ANGSD \citep{korneliussen2014angsd} and permits the exploration of genome-scale results through interactive visualization.
ANGSD-wrapper is under active development to incorporate new analyses and updates to the ANGSD software package.

\section*{Acknowledgments}
We acknowledge funding support from the US National Science Foundation (IOS-1238014 to JRI and IOS-1339393 to PLM), from a University of Minnesota Doctoral Dissertation Fellowship supporting TJYK, and from USDA Hatch project CA-D-PLS-2066-H, and from the University of California Davis Plant Sciences Department.  We thank members of the Ross-Ibarra and Morrell Labs for discussion and software testing. We thank the authors of ANGSD and related programs for answering questions, particularly Matteo Fumagalli and Filipe Vieira. Finally, we would like to thank Felix Andrews for statistical advice, although we did not follow it. 
\clearpage
\singlespacing
\bibliographystyle{apa}
\bibliography{references}
